### 2004-07-20 Documentation/Paper/paper.tex
Changes Meeting P+W made

\vspace*{.8cm}
%\vspace*{.3cm}
{\begin{center} {\large{\sc
                P.~Huber\footnote[1]{\makebox[1.cm]{Email:}
                phuber@ph.tum.de},~
                M.~Lindner\footnote[2]{\makebox[1.cm]{Email:}
                lindner@ph.tum.de},~and~
                W.~Winter\footnote[5]{\makebox[1.cm]{Email:}
                wwinter@ph.tum.de}
                }}
\end{center}}
\vspace*{0cm}
{\it
\begin{center}

\footnotemark[1]${}^,$\footnotemark[2]${}^,$\footnotemark[3]%
       Institut f\"ur theoretische Physik, Physik--Department,\\
       Technische Universit\"at M\"unchen,
       James--Franck--Strasse, D--85748 Garching, Germany

@@

\begin{abstract}
We present the \GLOBES\ (``General Long Baseline Experiment Simulator'')
software package, which allows the simulation of long-baseline and reactor
neutrino oscillation experiments. One part of the software is the
abstract experiment definition language to define experiments with
beam and full detector descriptions as accurate as possible. For the
definition of experiments, a new transparent building block concept is 
introduced. In addition, an additional program provides the possibility 
to develop and test new experiment definitions quickly.
%
Another
part of \GLOBES\ is the user's interface, which provides probability, rate, 
and $\Delta \chi^2$ information for a given experiment or any
combination of up to 32 experiments in C. Especially, the
$\Delta \chi^2$ functions allow a simulation with statistics only,
systematics, correlations, and degeneracies. In particular, \GLOBES\
can handle the full multi-parameter correlation among the oscillation
parameters, external input, and matter density uncertainties. Eventually,
many systematics options are provided, such as normalization and energy
calibration errors, or the choice between spectral or total rate 
information. 
\end{abstract}

@@

Neutrino oscillations are now established as the leading flavor
transition mechanism for neutrinos. In a long history of many experiments, see
\eg~\cite{Barger:2003qi}, 
two oscillation
frequencies have been identified: The fast ``atmospheric'' and the
slow ``solar'' oscillations, which are driven by the respective
mass squared differences.  In addition, there could be interference
effects between these two oscillations, provided that the coupling given by 
the small mixing angle $\theta_{13}$ is large enough. Such interference
effects include, for example, leptonic CP violation.  In order to
test the unknown oscillation parameters, \ie, the mixing angle $\theta_{13}$,
 the leptonic CP phase, and the
neutrino mass hierarchy, new long-baseline and reactor experiments 
are proposed.
These experiments send an artificial neutrino beam to a detector, or
detect the neutrinos produced by a nuclear fission reactor. However,
the presence of multiple solutions which are intrinsic to
 neutrino oscillation probabilities~\cite{Fogli:1996pv,
Burguet-Castell:2001ez,Minakata:2001qm,Barger:2001yr} 
affect these measurements.
Thus optimization strategies are required which maximally exploit 
complementarity between experiments. Therefore, a modern complete experiment 
simulation  and analysis tool does not only need to have 
a highly accurate beam and detector simulation but also powerful 
means to analyze correlations and degeneracies, especially for the combination
of several experiments.
In this paper, we present the \GLOBES\ software package
as such a tool.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=16cm]{GLOBES}
\end{center}
\caption{\label{fig:GLOBES} General concept of the \GLOBES\ package.}
\end{figure}

On the experiment definition side, either built-in neutrino fluxes
(\eg, neutrino factory) or arbitrary, user defined fluxes can be used. 
Similarly,
arbitrary cross sections, energy dependent efficiencies,
energy resolution functions as well as the considered oscillation channels 
backgrounds, and many other properties can be specified. 
For the systematics, energy
normalization and calibration errors can be simulated. Note that
energy ranges and windows and bin widths can be
(almost) arbitrarily chosen, including bins of different
widths. Together with \GLOBES\ comes a number of
pre-defined experiments in order to demonstrate the capabilities
of \GLOBES\ and to provide prototypes for new experiments.
Examples for these prototypes are the \MINOS , \ICARUS , and
\OPERA\ simulations from \Ref~\cite{Huber:2004ug}, the
\JHFSK\ and \NUMI\ superbeam simulations from \Refs~\cite{Huber:2002mx,
Huber:2002rs}, the \JHFHK\ superbeam upgrade simulation from
\Ref~\cite{Huber:2002mx}, the neutrino factory simulations from
\Ref~\cite{Huber:2002mx}, and the reactor experiment simulations from
\Ref~\cite{Huber:2003pm}. Other projects using the \GLOBES\ 
software include \Refs~\cite{Huber:2003ak,Ohlsson:2003ip,Winter:2003ye,Antusch:2004yx}.

@@

In this section, we sketch the computation of 
the event rates, which is on of the core parts of the \GLOBES\
software. However, because of the complexity of this issue,
we refer to the \GLOBES\ manual~\cite{Manual} for
details.

@@

Since it is a lot of effort to solve this double integral numerically,
we split up the two integrations. First, we evaluate the integral over
$\hat{E}$, where the only terms containing $\hat{E}$ are
$k_\beta^{\text{IT}}(E-\hat{E})$,  $ T_\beta(\hat{E})$, and 
$ V_\beta(\hat{E}-E')$. We define:
\begin{eqnarray}
\label{equ:e_res} 
R_\beta^{\text{IT}}(E,E')\,\epsilon_\beta^{\text{IT}}(E')
 \equiv
\int\limits_0^\infty d\hat{E} \quad T_\beta(\hat{E})\,k_\beta^{\text{IT}}(E-\hat{E})
\,V_\beta(\hat{E}-E')\,. 
\end{eqnarray}
Thus, $R_\beta^{\text{IT}}(E,E')$ describes the energy response of 
the detector, \ie , a neutrino with a (true) energy $E$ is reconstructed
with an energy between $E'$ and $E'+dE'$ with a probability
$R_\beta^{\text{IT}}(E,E') dE'$. The function $R(E,E')$ is also often called ``energy resolution function''. Actually, its internal representation
in the software is a smearing matrix. The function $\epsilon_\beta^{\text{IT}}(E')$ will be refered to as ``post-smearing efficiencies'', since it will allow us to define cuts and threshold functions {\em after} the smearing is performed, \ie, as function of $E'$. 
In addition, \GLOBES\ uses ``pre-smearing efficiencies'', which are 
evaluated {\em before} the smearing is performed, \ie, as function of $E$.
Similarly, constant\footnote{with respect to the oscillation parameters, 
not the energy} event rates can be added to the event rates before or 
after the energy smearing, which are refered to as 
``pre-smearing backgrounds'' and ``post-smearing background''. These types of 
efficiencies and backgrounds allow a very accurate modeling of
many factors, such as atmospheric or geoneutrino backgrounds, energy cuts, or
energy threshold functions.

@@

Core part of the event rate computation is the energy smearing algorithm
to evaluate \equ{events_bin}, where either a particular energy resolution
 function can be chosen for automated energy smearing, or the discretized
smearing matrix $R^c(E,E') = R^c_{ij} $ can be given manually. In addition, it is possible
to define a low-pass filter to avoid aliasing effects for neutrino
oscillations faster that the sampling width.

@@

The key concept of \AEDL\ is to regard a detector as an abstract system
which maps the true properties of a neutrino into the reconstructed properties
of the neutrino which subsequently are used in the fit of the oscillation
parameters. Within \GLOBES\ only the energy and flavour are used as 
observables, which is
sufficient for long baseline and reactor experiments. In  other
experiments, like \eg\ atmospheric experiments, more observables like
direction may have to be considered.

The main parts of any \AEDL\ experiment definition are channels,
rules, and experiments (\cf, \figu{aedl}). A channel corresponds to a neutrino oscillation channel including flux, cross section (for one specific interaction type), energy resolution function, initial and final neutrino flavors, their polarity (neutrinos or antineutrinos), and efficiencies. The channel raw event rates are computed according to \equ{events_bin}. Each channel leads to the raw event rates for a 
specific interaction type. In \AEDL , many of the channel components
have to be defined or loaded from files before, such as fluxes, 
cross sections, or the energy resolution function. For the
different options, see the \GLOBES\ manual~\cite{Manual}.

For a ``rule'', the raw event rates of one or more signal channels and
one or more background channels are added. The splitting in signal and
background is arbitrary, but all of the signal and background components
are defined to have a common systematics. The event rates of all 
signal and background components add up to the total event rate of the
rule, which leads to a $(\Delta \chi^2)_r$. The signal or background
within each rule allows the specification of signal and background 
normalization errors and energy tilt or calibration errors. 
These systematical errors are evaluated with the ``pull method'' like 
in~\cite{Huber:2002mx,Fogli:2002pt}.  
In addition to these
systematics errors, an overall evaluation strategy is assigned 
to each rule, which specifies the type of systematics (tilt or calibration
error), and the use of spectral information or total event rates.

Finally, one ore more rules add up to an experiment, where the
total $\Delta \chi^2$ is obtained as the sum of the $\Delta \chi^2$
of all rules. This approach allows the definition of 
appearance and disappearance channels, neutrino and antineutrino running, 
or interaction types with different systematics (spectral information
versus counting rate) within one experiment. The \GLOBES\ user interface
allows the simulation of one or more experiment simultaneously, which
means that one could also use different experiments for different
oscillation channels. However, there is still one component missing
in the experiment definition, which is the matter density profile.
For an experiment, an arbitrary matter density profiles can be specified, 
which is evaluated with the evolution operator method REFERENCES.
In addition, the matter density profile is multiplied by a scaling factor, 
which is treated as an independent parameter
with a relative (matter density) uncertainty. With this
approach, one can take into account that the matter density profile along a 
specific baseline is only known to about $3\%-5\%$.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=16cm]{projex}
\end{center}
\mycaption{\label{fig:projex} Left plot: Example for a  correlation between $\stheta$ and $\deltacp$ (for 1 d.o.f. only). Right plot: The $\chi^2$-value of the projection onto the $\stheta$-axis as function of $\stheta$. The projection onto the  $\stheta$-axis is obtained by finding the minimum $\chi^2$-value for each fixed value of $\stheta$ in the left-hand plot, \ie, along the gray vertical lines. The thick gray curve marks the position of these minima in the left-hand plot. The arrows mark the obtained fit ranges for $\stheta$ at the $3 \sigma$ confidence level (1 d.o.f.), \ie , the precision of $\stheta$.}
\end{figure}

In order to use \GLOBES , a C-library is linked with the application
software. This library provides the user's interface functions.
It allows to load \AEDL\ files, initialize the experiments, and
have access to various $\Delta \chi^2$-functions including any combination
of systematics and correlations. In addition, it provides low-level
access to oscillation probabilities and event rates, and allows
the readout and modification of many experiment parameters at running time.

The most sophisticated feature of the user's interface is the
possibility to include the full multi-parameter correlation among
all oscillation parameters. Together with the matter densities
of the $N$ loaded experiments ($N \ge 1$), the neutrino oscillation parameter space in \GLOBES\ has $6+N$ dimensions. In the $\chi^2$-approach,
the allowed region of $k$ ($1\le k<6+N$) fit parameters (at the chosen confidence level) is obtained as the projection of the $6+N$-dimensional fit manifold onto the $k$-dimensional hyperplane. For example,
for the precision of the parameter $\stheta$ only, the fit manifold
needs to be projected onto the $\stheta$-axis. One can easily imagine
that the computation of the full multi-parameter correlation is
very expensive with a grid-based method. However, the topology
of the neutrino oscillation parameter space is well investigated and
rather smooth. Therefore, it is feasible to use local minimization for each
solution in the parameter space.  For this process, 
a $6+N$-dimensional local minimizer
is started at the ``guessed'' position for the solution
(such as from analytical knowledge) and then runs into a
local minimum. This exercise has to be repeated till all local 
minima, \ie\ degenerate solutions, are found. In \figu{projex},
this minimization is illustrated at the simple example of a 
two-parameter correlation at a neutrino factory.

As we have discussed in the last section, the matter density of
each experiment is treated as an additional parameter, which is known
from external measurement with a certain precision. In principle, 
the \GLOBES\ user's interface allows such externally imposed
precisions for all oscillation parameters. This feature turns out
to be especially useful for input from the solar neutrino experiments,
since they provide better precisions for the solar neutrino
oscillation parameters $\sdm$ and $\theta_{12}$ than long-baseline
 experiments themselves.

@@

The running time of the systematics $(\Delta \chi^2)_{\mathrm{sys}}$ 
is 
\begin{equation}
\mathrm{t_{sys}} = \mathcal{O} \left( N_{\mathrm{Sampling}} \times N_{\mathrm{Bins}}  \, + \, N_{\mathrm{Sampling}} \times N_{\mathrm{Layers}} \right) ,
\end{equation}
since the oscillation probabilities for all channels and energies
are only computed once and stored in a list. Here $N_{\mathrm{Sampling}}$ is the number of sampling points for the evaluation of \equ{events_bin}, $N_{\mathrm{Bins}}$ is the number of energy bins,
and $N_{\mathrm{Layers}}$ is the number of matter density layers.\footnote{Note that we are here using the \AEDL\ format for these
parameters.}. Thus, whenever the
number of matter density steps is small, the product of sampling points
and bins dominates. In practice, the $N_{\mathrm{Sampling}} \times N_{\mathrm{Bins}}$ energy smearing matrix is already computed at 
the experiment initialization in order to save running time. In addition,
since it contains a lot of zeros, \GLOBES\ uses a special format to
avoid looping over the zeros.

The running time for the projected $\Delta \chi^2$ is, for a grid based method,
\begin{equation}
 \mathrm{t_{proj}^{Grid}} = \mathcal{O}\left( n^{6+N_{\mathrm{exp}}}\right) \, \times \, \mathrm{t_{sys}},
\end{equation}
where $n$ is the number of steps to be evaluated for each oscillation
parameter, and $N_{\mathrm{exp}}$ is the number of experiments
(since the matter densities of all experiments appear as parameters). 
Though it is, in principle, possible to use such a method with \GLOBES ,
it is impracticable in most cases, since it would involve at least 
billions of
steps for the complete multi-parameter correlation. The local $6+N_{\mathrm{exp}}$-dimensional minimization reduces this effort
to 
\begin{equation}
\mathrm{t_{proj}^{Minimizer}} = \mathcal{O}(1000) \, \times \,
N_{\mathrm{Deg}} \,  \times \, \mathrm{t_{sys}}
\end{equation}
steps, where $N_{\mathrm{Deg}}$ is the number of (disconnected) 
local minima, \ie, the number of
degeneracies.

@@

The installation of \GLOBES\ is highly automated by the use of the
Autotools family and follows the usual procedures. By default
a shared library {\tt libglobes.so} is installed but 
a static version is available
also. Additionally an executable with name {\tt globes} is installed. 
Provisions have been made to install \GLOBES\ without
root privilege into a user's home directory. During the automatic
configuration process also an example Makefile is produced which can
serve as a template for compiling and linking own applications against
{\tt libglobes.so}. This Makefile can be used to directly compile and
link the examples
from the manual.

@@

In summary, the \GLOBES\ software package provides powerful tools
to do a full three-flavor analysis of future long-baseline and reactor
 experiments
including systematics, correlations, and degeneracies. In addition,
the abstract experiment definition language allows to define
experiments at an abstract level in a highly accurate fashion. 
Some of the major strengths of \GLOBES\ are the ability to quickly
define new experiments, the potential to take into account the
full multi-parameter correlation, the possibilities to include
external input and matter density uncertainties and last but not least
the ability to jointly analyze several experiments.

@@

For the future it is planed to further extend the flexibility of
\GLOBES\ by providing support for dynamically loadable modules, to
allow completely user defined systematics or non-standard oscillation
probabilities.

@@

We would like to thank Martin Freund, who wrote the very first
version of a three-flavour matter profile treatment many years ago.
PH is especially thankful for the invaluable advice of Thomas Fischbacher on
many design issues. 
In addition, we would like to thank
Mark Rolinec for his help to translate the experiment descriptions
into \AEDL , and for creating the illustrations in this manuscript.
Finally, thanks to all the people who have been pushing this project
for many years, to the ones who have been continuing asking for the 
publication of the software, and the referees of several of our
papers for suggestions which lead to improvements in the software.
Parts of this works have been supported by the ``Sonderforschungsbereich
375 f\"ur Astro-Teilchenphysik der Deutschen Forschungsgemeinschaft''.

>>>>

\vspace*{.8cm}
%\vspace*{.3cm}
{\begin{center} {\large{\sc
                P.~Huber\footnote[1]{\makebox[1.cm]{Email:}
                phuber@ph.tum.de},~
                M.~Lindner\footnote[2]{\makebox[1.cm]{Email:}
                lindner@ph.tum.de},~and~
                W.~Winter\footnote[3]{\makebox[1.cm]{Email:}
                wwinter@ph.tum.de}
                }}
\end{center}}
\vspace*{0cm}
{\it
\begin{center}

\footnotemark[1]${}^,$\footnotemark[2]${}^,$\footnotemark[3]%
       Institut f\"ur Theoretische Physik, Physik--Department,\\
       Technische Universit\"at M\"unchen,
       James--Franck--Strasse, D--85748 Garching, Germany

@@

\begin{abstract}
We present the \GLOBES\ (``General Long Baseline Experiment Simulator'')
software package, which allows the simulation of long-baseline and reactor
neutrino oscillation experiments. One part of the software is the
abstract experiment definition language to define experiments with
beam and full detector descriptions as accurate as possible. Many systematics options are provided, such as normalization and energy
calibration errors, or the choice between spectral or total rate 
information. For the 
definition of experiments, a new transparent building block concept is 
introduced. In addition, an additional program provides the possibility 
to develop and test new experiment definitions quickly.
%
Another
part of \GLOBES\ is the user's interface, which provides probability, rate, 
and $\Delta \chi^2$ information for a given experiment or any
combination of up to 32 experiments in C. Especially, the
$\Delta \chi^2$ functions allow a simulation with statistics only,
systematics, correlations, and degeneracies. In particular, \GLOBES\
can handle the full multi-parameter correlation among the oscillation
parameters, external input, and matter density uncertainties. 
\end{abstract}

@@

Neutrino oscillations are now established as the leading flavor
transition mechanism for neutrinos. In a long history of many experiments, see
\eg~\cite{Barger:2003qi}, 
two oscillation
frequencies have been identified: The fast ``atmospheric'' and the
slow ``solar'' oscillations, which are driven by the respective
mass squared differences.  In addition, there could be interference
effects between these two oscillations, provided that the coupling given by 
the small mixing angle $\theta_{13}$ is large enough. Such interference
effects include, for example, leptonic CP violation.  In order to
test the unknown oscillation parameters, \ie, the mixing angle $\theta_{13}$,
 the leptonic CP phase, and the
neutrino mass hierarchy, new long-baseline and reactor experiments 
are proposed.
These experiments send an artificial neutrino beam to a detector, or
detect the neutrinos produced by a nuclear fission reactor. However,
the presence of multiple solutions which are intrinsic to
 neutrino oscillation probabilities~\cite{Fogli:1996pv,
Burguet-Castell:2001ez,Minakata:2001qm,Barger:2001yr} 
affect these measurements.
Thus optimization strategies are required which maximally exploit 
complementarity between experiments. Therefore, a modern, complete experiment 
simulation  and analysis tool does not only need to have 
a highly accurate beam and detector simulation, but also powerful 
means to analyze correlations and degeneracies, especially for the combination
of several experiments.
In this paper, we present the \GLOBES\ software package
as such a tool.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=14cm]{GLOBES}
\end{center}
\caption{\label{fig:GLOBES} General concept of the \GLOBES\ package.}
\end{figure}

On the experiment definition side, either built-in neutrino fluxes
(\eg, neutrino factory) or arbitrary, user defined fluxes can be used. 
Similarly,
arbitrary cross sections, energy dependent efficiencies,
energy resolution functions as well as the considered oscillation channels, 
backgrounds, and many other properties can be specified. 
For the systematics, energy
normalization and calibration errors can be simulated. Note that
energy ranges and windows and bin widths can be
(almost) arbitrarily chosen, including bins of different
widths. Together with \GLOBES\ comes a number of
pre-defined experiments in order to demonstrate the capabilities
of \GLOBES\ and to provide prototypes for new experiments.
In addition, they can be used to test new physics ideas with
complete experiment simulations.
Examples for these prototypes are the \MINOS , \ICARUS , and
\OPERA\ simulations from \Ref~\cite{Huber:2004ug}, the
\JHFSK\ and \NUMI\ superbeam simulations from \Refs~\cite{Huber:2002mx,
Huber:2002rs}, the \JHFHK\ superbeam upgrade simulation from
\Ref~\cite{Huber:2002mx}, the neutrino factory simulations from
\Ref~\cite{Huber:2002mx}, and the reactor experiment simulations from
\Ref~\cite{Huber:2003pm}. Other projects using earlier versions
of the \GLOBES\ software include \Refs~\cite{Huber:2003ak,Ohlsson:2003ip,Winter:2003ye,Antusch:2004yx}.

@@

In this section, we sketch the computation of 
the event rates, which is one of the core parts of the \GLOBES\
software. However, because of the complexity of this issue,
we refer to the \GLOBES\ manual~\cite{Manual} for
details.

@@

Since it is a lot of effort to solve this double integral numerically,
we split up the two integrations. First, we evaluate the integral over
$\hat{E}$, where the only terms containing $\hat{E}$ are
$k_\beta^{\text{IT}}(E-\hat{E})$,  $ T_\beta(\hat{E})$, and 
$ V_\beta(\hat{E}-E')$. We define:
\begin{eqnarray}
\label{equ:e_res} 
R_\beta^{\text{IT}}(E,E')\,\epsilon_\beta^{\text{IT}}(E')
 \equiv
\int\limits_0^\infty d\hat{E} \quad T_\beta(\hat{E})\,k_\beta^{\text{IT}}(E-\hat{E})
\,V_\beta(\hat{E}-E')\,. 
\end{eqnarray}
Thus, $R_\beta^{\text{IT}}(E,E')$ describes the energy response of 
the detector, \ie , a neutrino with a (true) energy $E$ is reconstructed
with an energy between $E'$ and $E'+dE'$ with a probability
$R_\beta^{\text{IT}}(E,E') dE'$. The function $R(E,E')$ is also often called ``energy resolution function''. Actually, its internal representation
in the software is a smearing matrix. The function $\epsilon_\beta^{\text{IT}}(E')$ will be refered to as ``post-smearing efficiencies'', since it will allow us to define cuts and threshold functions {\em after} the smearing is performed, \ie, as function of $E'$. 
In addition, \GLOBES\ uses ``pre-smearing efficiencies'', which are 
evaluated {\em before} the smearing is performed, \ie, as function of $E$.
Similarly, constant\footnote{With respect to the oscillation parameters, 
not the energy.} background rates can be added to the event rates before or 
after the energy smearing, which are refered to as 
``pre-smearing backgrounds'' and ``post-smearing backgrounds''. These types of 
efficiencies and backgrounds allow a very accurate modeling of
many factors, such as atmospheric or geoneutrino backgrounds, energy cuts, or
energy threshold functions.

@@

Core part of the event rate computation is the energy smearing algorithm
to evaluate \equ{events_bin}, where either a particular energy resolution
 function can be chosen for automated energy smearing, or the discretized
smearing matrix $R^c(E,E') = R^c_{ij} $ can be given manually. In addition, it is possible
to define a low-pass filter to avoid aliasing effects for neutrino
oscillations faster than the sampling width.

@@

The key concept of \AEDL\ is to regard a detector as an abstract system
which maps the true properties of a neutrino onto the reconstructed properties
of the neutrino. The latter are subsequently used in the fit of the oscillation
parameters. Within \GLOBES , only energy and flavor are  
observables, which is
sufficient for long-baseline and reactor experiments. In  other
experiments, such as atmospheric neutrino experiments, more observables 
(\eg , direction) may have to be considered.

The main components of any \AEDL\ experiment definition are ``channels'',
``rules'', and ``experiments'' (\cf, \figu{aedl}). A channel corresponds to a neutrino oscillation channel including flux, cross section (for one specific interaction type), energy resolution function, initial and final neutrino flavors, their polarity (neutrinos or antineutrinos), and efficiencies. The channel raw event rates are computed according to \equ{events_bin}. Each channel leads to the raw event rates for a 
specific interaction type. In \AEDL , many of the channel components
have to be defined or loaded from files before, such as fluxes, 
cross sections, or the energy resolution function. For the
different options, see the \GLOBES\ manual~\cite{Manual}.

For a ``rule'', the raw event rates of one or more signal channels and
one or more background channels are added. The splitting in signal and
background is arbitrary, but all of the signal and background components
are defined to have a common systematics. The event rates of all 
signal and background components add up to the total event rate of the
rule, which leads to a $(\Delta \chi^2)_r$. The signal or background
within each rule allows the specification of signal and background 
normalization errors and energy tilt or calibration errors. 
These systematical errors are evaluated with the ``pull method'' such as 
in \Refs~\cite{Huber:2002mx,Fogli:2002pt}.  
In addition to these
systematics errors, an overall evaluation strategy is assigned 
to each rule, which specifies the type of systematics (tilt or calibration
error), and the use of spectral information or total event rates.

Finally, one ore more rules add up to an experiment, where the
total $\Delta \chi^2$ is obtained as the sum of the $(\Delta \chi^2)_r$
of all rules. This approach allows the definition of 
appearance and disappearance channels, neutrino and antineutrino running, 
or interaction types with different systematics (spectral information
versus counting rate) within one experiment. The \GLOBES\ user interface
allows the simulation of one or more experiment simultaneously, which
means that one could also use different experiments for different
oscillation channels. However, there is still one component missing
in the experiment definition, which is the matter density profile.
For an experiment, an arbitrary matter density profile can be specified, 
which is evaluated with the evolution operator method~\cite{Ohlsson:1999um}.
In addition, the matter density profile is multiplied by a scaling factor, 
which is treated as an independent parameter
with a relative (matter density) uncertainty. With this
approach, one can take into account that the matter density profile along a 
specific baseline is only known to about $3\%-5\%$. Thus, for one
particular baseline, all rules should be defined within one experiment.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=16cm]{projex}
\end{center}
\mycaption{\label{fig:projex} Left plot: Example for a  correlation between $\stheta$ and $\deltacp$ (for 1 d.o.f. only, un-shown oscillation
parameters fixed). Right plot: The $\chi^2$-value of the projection onto the $\stheta$-axis as function of $\stheta$. The projection onto the  $\stheta$-axis is obtained by finding the minimum $\chi^2$-value for each fixed value of $\stheta$ in the left-hand plot, \ie, along the gray vertical lines. The thick gray curve marks the position of these minima in the left-hand plot. The arrows mark the obtained fit ranges for $\stheta$ at the $3 \sigma$ confidence level (1 d.o.f.), \ie , the precision of $\stheta$.}
\end{figure}

In order to use \GLOBES , a C-library is linked with the application
software. This library provides the user's interface functions.
It allows to load \AEDL\ files, initialize the experiments, and
have access to various $\Delta \chi^2$-functions including any combination
of systematics and correlations. In addition, it provides low-level
access to oscillation probabilities and event rates, and allows
the readout and modification of many experiment parameters at run time.

The most sophisticated feature of the user's interface is the
possibility to include the full multi-parameter correlation among
all oscillation parameters. Together with the matter densities
of the $N_{\mathrm{exp}}$ loaded experiments ($N_{\mathrm{exp}} \ge 1$), the neutrino oscillation parameter space in \GLOBES\ has $6+N_{\mathrm{exp}}$ dimensions. In the $\chi^2$-approach,
the allowed region of $k$ ($1\le k<6+N_{\mathrm{exp}}$) fit parameters (at the chosen confidence level) is obtained as the projection of the $6+N_{\mathrm{exp}}$-dimensional fit manifold onto the $k$-dimensional hyperplane. For example,
for the precision of the parameter $\stheta$ only, the fit manifold
needs to be projected onto the $\stheta$-axis. One can easily imagine
that the computation of the full multi-parameter correlation is
very expensive with a grid-based method. However, the topology
of the neutrino oscillation parameter space is well investigated and
rather smooth. Therefore, it is feasible to use local minimization for each
solution in the parameter space.  For this process, 
a $6+N_{\mathrm{exp}}$-dimensional local minimizer
is started at the ``guessed'' position for the solution
(such as from analytical knowledge) and then runs into a
local minimum. This exercise has to be repeated till all local 
minima, \ie\ degenerate solutions, are found. In \figu{projex},
this minimization is illustrated at the simple example of a 
two-parameter correlation at a neutrino factory.

As we have indicated above, the matter density of
each experiment is treated as an additional parameter, which is known
from external measurement with a certain precision. In principle, 
the \GLOBES\ user's interface allows such externally imposed
precisions for all oscillation parameters. This feature turns out
to be especially useful for input from the solar neutrino experiments,
since they provide better precisions for the solar neutrino
oscillation parameters $\sdm$ and $\theta_{12}$ than long-baseline
 experiments themselves.

@@

The run time of the systematics $(\Delta \chi^2)_{\mathrm{sys}}$ 
is 
\begin{equation}
\mathrm{t_{sys}} = \mathcal{O} \left( N_{\mathrm{Sampling}} \times N_{\mathrm{Bins}}  \, + \, N_{\mathrm{Sampling}} \times N_{\mathrm{Layers}} \right) ,
\end{equation}
since the oscillation probabilities for all channels and energies
are only computed once and stored in a list. Here $N_{\mathrm{Sampling}}$ is the number of sampling points for the evaluation of the right integral in \equ{events_bin}, $N_{\mathrm{Bins}}$ is the number of energy bins,
and $N_{\mathrm{Layers}}$ is the number of matter density layers. Thus, whenever the
number of matter density steps is small, the product of sampling points
and bins dominates. In practice, the $N_{\mathrm{Sampling}} \times N_{\mathrm{Bins}}$ energy smearing matrix is already computed at 
the experiment initialization in order to save run time. In addition,
since it contains a lot of zeros, \GLOBES\ uses a special format to
avoid looping over the zeros.

The run time for the projected $\Delta \chi^2$ would, for a grid based method, be
\begin{equation}
 \mathrm{t_{proj}^{Grid}} = \mathcal{O}\left( n^{6+N_{\mathrm{exp}}}\right) \, \times \, \mathrm{t_{sys}},
\end{equation}
where $n$ is the number of grid points to be evaluated for each oscillation
parameter, and $N_{\mathrm{exp}}$ is the number of experiments
(since the matter densities of all experiments appear as parameters). 
Though it is, in principle, possible to use such a method with \GLOBES ,
it is impracticable in most cases, since it would involve at least 
billions of
steps for the complete multi-parameter correlation. The local $6+N_{\mathrm{exp}}$-dimensional minimization reduces this effort
to 
\begin{equation}
\mathrm{t_{proj}^{Minimizer}} = \mathcal{O}(1000) \, \times \,
N_{\mathrm{Deg}} \,  \times \, \mathrm{t_{sys}}
\end{equation}
steps, where $N_{\mathrm{Deg}}$ is the number of (disconnected) 
local minima, \ie, the number of
degeneracies (typically $\mathcal{O}(10)$).

@@

The installation of \GLOBES\ is highly automated by the use of the
Autotools family and follows the usual procedures. By default
a shared library {\tt libglobes.so} is installed, but 
a static version is available, too. In addition, an executable named {\tt globes} is installed. It is another possibility to install \GLOBES\ without
root privilege into a user's home directory. During the automatic
configuration process also an example Makefile is produced which can
serve as a template for compiling and linking own applications against
{\tt libglobes.so}. This Makefile can be used to directly compile and
link the examples
from the manual.

@@

In summary, the \GLOBES\ software package provides powerful tools
to do a full three-flavor analysis of future long-baseline and reactor
 experiments
including systematics, correlations, and degeneracies. In addition,
the abstract experiment definition language allows to define
experiments at an abstract level in a highly accurate fashion. 
Some of the major strengths of \GLOBES\ are the ability to quickly
define new experiments, the potential to take into account the
full multi-parameter correlation, the possibilities to include
external input and matter density uncertainties and 
the ability to analyze several experiments simultaneously.

@@

@@

We would like to thank Martin Freund, who wrote the very first
version of a three-flavor matter profile treatment many years ago.
PH is especially thankful for the invaluable advice of Thomas Fischbacher on
many design issues. 
In addition, we would like to thank
Mark Rolinec for his help to translate the experiment descriptions
into \AEDL , and for creating the illustrations in this manuscript.
Finally, thanks to all the people who have been pushing this project
for many years, to the ones who have been continuing asking for the 
publication of the software, and the referees of several of our
papers for suggestions which lead to improvements in the software.
This work has been supported by the ``Sonderforschungsbereich
375 f\"ur Astro-Teilchenphysik der Deutschen Forschungsgemeinschaft''.
